PyPy's Approach to Virtual Machine Construction

### 2006-05-31 talk/dls2006/paper.tex
typos, wording tweaks, small fixes.  i'm done, i think.

We shortly describe the architecture of PyPy in section
\ref{architecture}.  In section \ref{systemprog} we describe our
approach of varying the type systems at various levels of the
translation.  Section \ref{typeinference} gives an overview of the
type inference engine we developed (and can be read independently from
section 3.)  We present experimental results in section
\ref{experimentalresults} and future work directions in section
\ref{futurework}.  In section \ref{relatedwork} we compare with
related work, and finally we conclude in section \ref{conclusion}.

@@

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environment.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough types to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code, it is the largest RPython
program that we have translated so far.  More information can be found
in \cite{A}.

@@

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's innards.  It can inspect all the graphs to discover the
\texttt{struct} types in use by the program, and assign a unique type id to
each of them.  These type ids are collected in internal tables that
describe the layout of the structures, e.g.\ their sizes and the location
of the pointer fields.

@@

In the example of the \texttt{malloc} operation, replaced by a call to
GC code, this GC code can invoke a complete collection of dead
objects, and can thus be arbitrarily complicated.  Still, our GC code
is entirely written in plain Python, and it manipulates ``objects'' that
are still at a lower level: pointer and address objects.  Even with
the restriction of having to use pointer-like and address-like
objects, Python remains more expressive than, say, C to write a GC
(Jikes RVM's GC work \cite{JikesGC} was the inspiration to try to
express GCs in Python, see section \ref{relatedwork}).

@@

The RPython level is a subset of Python, so the types mostly follow
Python types, and the instances of these types are instances in the
normal Python sense; e.g.\ whereas Python has only got a single type
\texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
type \texttt{T}, but instances of \texttt{list(T)} are just those Python lists
whose items are all instances of \texttt{T}.

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:
%
\begin{enumerate}
\item the types, which we use to tag the variables of the graphs at
      the given level.  (Types are actually annotated, self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly.)

\item the Python objects that emulate instances of these types.  (More
      about them below.)
\end{enumerate}
%
We have defined well-typed operations between instances of these types,
syntactically expressed with standard Python operators (e.g.\ if \texttt{x}
is a C-level array, \texttt{x[n]} accesses its \texttt{n}th item).
The emulating instances provide a concrete implementation of
these operations that works in normal Python; the types involved in the
operations are also known to the type inference engine when it
analyses system code like the helpers of figure \ref{fig_llappend}.

Now, clearly, the purpose of types like a ``C-like struct'' or a ``C-like
array'' is to be translated to a real \texttt{struct} or array declaration by
the C back-end.  What, then, is the purpose of emulating such things in
Python?  The answer is three-fold.  Firstly, if we have objects that
live within the Python interpreter, but faithfully emulate the behavior
of their C equivalent while performing additional safety checks, they
are an invaluable help for testing and debugging.  For example, we can
check the correctness of our hash table implementation, written in
Python in term of struct- and array-like objects, just by running it.
The same holds for the GC.

Secondly, and anecdotically, as the type inference process (section
\ref{typeinference}) is based on abstract interpretation, we can use
the following trick: the resulting type of most low-level operations
is deduced simply by example.  Sample C-level objects are
instantiated, used as arguments to a given operation, and produce a
sample result, whose C-level type must be the type of the result
variable in the graph.

@@

The various analyses used -- from type inference to lifetime analysis -
are generally formulated as [abstract interpretation].  While this
approach is known to be less efficient than more tailored algorithms
like constraint-based type inference, we gain in freedom,
controllability and simplicity.  This proved essential in our overall
approach: as described in section \ref{systemprog}, we need to perform
type inference with many different type systems, the details of which
have evolved along the road.

@@

This order is extremely simple, because most actual analysis is delayed
to the next phase, the type inference engine.  The objects are either
\textit{Variables}, which are pure placeholders for entierely unknown values,
or \textit{Constants} with a concrete Python object as value.  The order places
Variable as the top, and keeps all \textit{Constants} unordered.  Thus if two
different constants merge during abstract interpretation, we immediately
widen them to Variable.

@@

Note that we produce flow graphs in Static Single Information (SSI, \cite{SSI})
form, an extension of Static Single Assignment (\cite{SSA}): each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed.

While the Flow Object Space is quite a short piece of code -- its core
functionality holds in 300 lines -- the detail of the interactions
sketched above is not entierely straightforward; we refer the reader to
\cite{D} for more information.

@@

The goal of the annotator is to find the most precise type that can be
given to each variable of all control flow graphs while respecting the
constraints imposed by the operations in which these variables are
involved.

@@

We can consider that all variables are initially assigned the ``bottom''
type corresponding to the empty set of possible run-time values.  Types
can only ever be generalised, and the model is simple enough to show
that there is no infinite chain of generalisation, so that this process
necessarily terminates.

@@

As seen in section \ref{systemprog}, we use the annotator with more than one type
systems.  The more interesting and complex one is the RPython type
system, which describes how the input RPython program can be annotated.
The other type systems contain lower-level, C-like types that are mostly
unordered, thus forming more trivial lattices than the one formed by
RPython types.

@@

All list terms for all variables are unordered.  The Pbcs form a
classical finite set-of-subsets lattice.  In addition, we have left
out a number of other annotations that are irrelevant for the basic
description of the annotator and straightforward to handle:
$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
complete list is described in document \cite{T}.

@@

The worst case behaviors that can appear in the model described above
involve the lattice of Pbcs, involving variables that could contain
e.g.\ one function object among many.  An example of such behavior is
code manipulating a table of function objects: when an item is read
out of the table, its type is a large Pbc set: $Pbc(\{f1, f2, f3,
\ldots\})$.  But in this example, the whole set is available at once,
and not built incrementally by successive discoveries.  This seems to
be often the case in practice: it is not very common for programs to
manipulate objects that belong to a large but finite family -- and when
they do, the whole family tends to be available early on, requiring
few reflowing.

@@

Our tool-chain is capable of translating the Python interpreter of
PyPy, written in RPython, producing right now either ANSI C code as
described before, or LLVM\footnote{The LLVM project is the realisation
of a portable assembler infrastructure, offering both a virtual
machine with JIT capabilities and static compilation. Currently we are
using the latter with its good high-level optimisations for PyPy.}
assembler which is then natively compiled with LLVM tools.

@@

{\bf pypy-c.}
    The simplest variant: translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC.

{\bf pypy-c-thread.}
    The same, with OS thread support enabled.  (For measurement purposes,
    thread support is kept separate because it has an impact on the GC
    performance.)

{\bf pypy-c-stackless.}
    The same as pypy-c, plus the ``stackless transformation'' step which
    modifies the flow graph of all functions in a way that allows them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework.}
    In this variant, the ``gc transformation'' step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework.}
    This variant combines the ``gc transformation'' step with the
    ``stackless transformation'' step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

{\bf pypy-llvm-c.}
    The same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

{\bf pypy-llvm-c-prof.}
    The same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than more customized approach;
kernel-level profiling shows that pypy-c typically spends 30\% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC is even quite worse.  The interaction with processor caches is also
hard to predict and account for; in general, we tend to produce
relatively large amounts of code and prebuilt data.

@@

To achieve high performance for dynamic languages such as Python, the
proven approach is to use dynamic compilation techniques, i.e.\ to write
JITs.  With direct techniques, this is however a major endeavour, and
increases the efforts to further evolve the language.

In the context of the PyPy project, we are now exploring -- as we planned
from the start -- the possibility to produce a JIT as a graph
transformation aspect from the Python interpreter.  This idea is based
on the theoretical possibiliy to turn interpreters into compilers by
partial evaluation\cite{Jones:1993:PartialEvaluation}.
In our approach, this is done by analysing
the forest of flow graphs built from the Python interpreter, which is a
well-suited input for this kind of techniques.  We can currently perform
binding-time analysis on these graphs, again with abstract
interpretation techniques reusing the type inference engine.  The next
step is to transform the graphs -- following the binding-time annotations
- into a compiler; more precisely, in partial evalution terminology, a
generating extension.  We can currently do this on trivial examples.

@@

Jikes RVM's native JIT compilers \cite{Jikes-JIT}
are not meant to be retargetted to run in other environments
that hardware processors, for example in a CLR/.NET
runtime. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code.  Major changes of the
language are likely to correspond to major surgery of the JIT.

@@

Targetting platforms that are very different from C/Posix is work in
progress, but given that many of the initial components are shared with
the existing stack of transformations leading to C, we are confident
that this work will soon give results.  Moreover, we believe that these
results will show a reasonable efficiency, because the back-ends for VMs
like Squeak and .NET can take advantage of high-level input (as opposed
to trying to translate, say, C-like code to Smalltalk).

>>>>

We shortly describe the architecture of PyPy in section
\ref{architecture}.  In section \ref{systemprog} we describe our
approach of varying the type systems at various levels of the
translation.  Section \ref{typeinference} gives an overview of the
type inference engine we developed (and can be read independently from
section 3).  We present experimental results in section
\ref{experimentalresults} and future work directions in section
\ref{futurework}.  In section \ref{relatedwork} we compare with
related work, and finally we conclude in section \ref{conclusion}.

@@

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environments.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough type information to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than to mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code, it is the largest RPython
program that we have translated so far.  More information can be found
in \cite{A}.

@@

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's allocation routine.  It can inspect all the graphs to discover the
\texttt{struct} types in use by the program, and assign a unique type id to
each of them.  These type ids are collected in internal tables that
describe the layout of the structures, e.g.\ their sizes and the location
of the pointer fields.

@@

In the example of the \texttt{malloc} operation, replaced by a call to
GC code, this GC code can invoke a complete collection of dead
objects, and can thus be arbitrarily complicated.  Still, our GC code
is entirely written in plain Python, and it manipulates ``objects'' that
are still at a lower level: pointer and address objects.  Even with
the restriction of having to use pointer-like and address-like
objects, Python remains more expressive than, say, C to write a GC
(the work on the Jikes RVM's GC \cite{JikesGC} was the inspiration to try to
express GCs in Python, see section \ref{relatedwork}).

@@

The RPython level is a subset of Python, so the types mostly follow
Python types, and the instances of these types are instances in the
normal Python sense; for example where Python has only a single type
\texttt{list}, RPython has a parametric type \texttt{list(T)} for every RPython
type \texttt{T}, but instances of \texttt{list(T)} are just those Python lists
whose items are all instances of \texttt{T}.

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:
%
\begin{enumerate}
\item The types, which we use to tag the variables of the graphs at
      the given level (types are actually annotated, self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly).

\item The Python objects that emulate instances of these types (more
      about them below).
\end{enumerate}
%
We have defined well-typed operations between instances of these types,
syntactically expressed with standard Python operators (e.g.\ if \texttt{x}
is a C-level array, \texttt{x[n]} accesses its \texttt{n}th item).
The emulating instances provide a concrete implementation of
these operations that works in normal Python; the types involved in the
operations are also known to the type inference engine when it
analyses system code like the helpers of figure \ref{fig_llappend}.

Now, clearly, the purpose of types like a ``C-like struct'' or a ``C-like
array'' is to be translated to a real \texttt{struct} or array declaration by
the C back-end.  What, then, is the purpose of emulating such things in
Python?  The answer is three-fold.  Firstly, having objects that
live within the Python interpreter, but faithfully emulate the behavior
of their C equivalent while performing additional safety checks, is 
an invaluable help for testing and debugging.  For example, we can
check the correctness of our hash table implementation, written in
Python in term of struct- and array-like objects, just by running it.
The same holds for the GC.

Secondly, and anecdotally, as the type inference process (section
\ref{typeinference}) is based on abstract interpretation, we can use
the following trick: the resulting type of most low-level operations
is deduced simply by example.  Sample C-level objects are
instantiated, used as arguments to a given operation, and produce a
sample result, whose C-level type must be the type of the result
variable in the graph.

@@

The various analyses used -- from type inference to lifetime analysis -
are generally formulated as [abstract interpretation].  While this
approach is known to be less efficient than more tailored algorithms
like constraint-based type inference, we gain in freedom,
controllability and simplicity.  This proved essential in our overall
approach: as described in section \ref{systemprog}, we need to perform
type inference with many different type systems, the details of which
have evolved over time.

@@

This order is extremely simple, because most actual analysis is delayed
to the next phase, the type inference engine.  The objects are either
\textit{Variables}, which are pure placeholders for entirely unknown values,
or \textit{Constants} with a concrete Python object as value.  The order places
Variable as the top, and keeps all \textit{Constants} unordered.  Thus if two
different constants merge during abstract interpretation, we immediately
widen them to Variable.

@@

Note that we produce flow graphs in Static Single Information or SSI\cite{SSI}
form, an extension of Static Single Assignment\cite{SSA}: each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed.

While the Flow Object Space is quite a short piece of code -- its core
functionality takes only 300 lines -- the detail of the interactions
sketched above is not entirely straightforward; we refer the reader to
\cite{D} for more information.

@@

The goal of the annotator is to find the most precise of our types that can be
given to each variable of all control flow graphs while respecting the
constraints imposed by the operations in which these variables are
involved.

@@

We can consider that all variables are initially assigned the ``bottom''
type corresponding to the empty set of possible run-time values.  Types
can only ever be generalised, and the model is simple enough to show
that there is no infinite chain of generalization, so that this process
necessarily terminates.

@@

As seen in section \ref{systemprog}, we use the annotator with more than one type
systems.  The most interesting and complex one is the RPython type
system, which describes how the input RPython program can be annotated.
The other type systems contain lower-level, C-like types that are mostly
unordered, thus forming more trivial lattices than the one formed by
RPython types.

@@

All list terms for all variables are unordered.  The Pbcs form a
classical finite set-of-subsets lattice.  In addition, we have left
out a number of other annotations that are irrelevant for the basic
description of the annotator and straightforward to handle:
$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
complete list is described in \cite{T}.

@@

The worst case behaviors that can appear in the model described above
involve the lattice of Pbcs, involving variables that could contain
e.g.\ one function object among many.  An example of such behavior is
code manipulating a table of function objects: when an item is read
out of the table, its type is a large Pbc set: $Pbc(\{f_1, f_2, f_3,
\ldots\})$.  But in this example, the whole set is available at once,
and not built incrementally by successive discoveries.  This seems to
be often the case in practice: it is not very common for programs to
manipulate objects that belong to a large but finite family -- and when
they do, the whole family tends to be available early on, requiring
few reflowing.

@@

Our tool-chain is capable of translating the Python interpreter of
PyPy, written in RPython, currently producing either ANSI C code as
described before, or LLVM\footnote{The LLVM project is the realization
of a portable assembler infrastructure, offering both a virtual
machine with JIT capabilities and static compilation. Currently we are
using the latter with its good high-level optimizations for PyPy.}
assembler which is then compiled to native code with LLVM tools.

@@

{\bf pypy-c:}
    The simplest variant; translated to C code with no explicit memory
    management, and linked with the Boehm conservative GC.

{\bf pypy-c-thread:}
    The same, with OS thread support enabled (thread support is kept
    separate for measurement purposes because it has an impact on the
    GC performance).

{\bf pypy-c-stackless:}
    The same as pypy-c, plus the ``stackless transformation'' step which
    modifies the flow graph of all functions to allow them
    to save and restore their local state, as a way to enable coroutines.

{\bf pypy-c-gcframework:}
    In this variant, the ``gc transformation'' step inserts explicit
    memory management and a simple mark-and-sweep GC implementation.
    The resulting program is not linked with Boehm.  Note that it is not
    possible to find all roots from the C stack in portable C; instead,
    in this variant each function explicitly pushes and pops all roots
    to an alternate stack around each subcall.

{\bf pypy-c-stackless-gcframework:}
    This variant combines the ``gc transformation'' step with the
    ``stackless transformation'' step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

{\bf pypy-llvm-c:}
    The same as pypy-c, but using the LLVM back-end instead of the C
    back-end.  The LLVM assembler-compiler gives the best results when -
    as we do here -- it optimizes its input and generates again C code,
    which is fed to GCC.

{\bf pypy-llvm-c-prof:}
    The same as pypy-llvm-c, but using GCC's profile-driven
    optimizations.

The speed difference with CPython 2.4.3 can be explained at two levels.
One is that CPython is hand-crafted C code that has been continuously
optimized for a decade now, whereas the Python interpreter of PyPy first
seeks flexibility and high abstraction levels.  The other, probably
dominant, factor is that various indices show that our approach places a
very high load on the GC and on the memory caches of the machine.  The
Boehm GC is known to be less efficient than more customized approach;
kernel-level profiling shows that pypy-c typically spends 30\% of its
time in the Boehm library.  Our current, naively simple mark-and-sweep
GC is manages to be a bit worse.  The interaction with processor caches is also
hard to predict and account for; in general, we tend to produce
relatively large amounts of code and prebuilt data.

@@

To achieve high performance for dynamic languages such as Python, the
proven approach is to use dynamic compilation techniques, i.e.\ to write
JITs.  With direct techniques, this is however a major endeavour, and
increases the effort involved in further evolution of the language.

In the context of the PyPy project, we are now exploring -- as we planned
from the start -- the possibility of producing a JIT as a graph
transformation aspect from the Python interpreter.  This idea is based
on the theoretical possibiliy to turn interpreters into compilers by
partial evaluation\cite{Jones:1993:PartialEvaluation}.
In our approach, this is done by analysing
the forest of flow graphs built from the Python interpreter, which is a
well-suited input for this kind of techniques.  We can currently perform
binding-time analysis on these graphs, again with abstract
interpretation techniques reusing the type inference engine.  The next
step is to transform the graphs -- following the binding-time annotations
-- into a compiler; more precisely, in partial evalution terminology, a
generating extension.  We can currently do this on trivial examples.

@@

Jikes RVM's native JIT compilers \cite{Jikes-JIT}
are not meant to be retargetted to run in other environments
than hardware processors, for example in a CLR/.NET
runtime. Also Jikes RVM pays the complexity of writing
a JIT up-front, which also means that features and semantics of the
language are encoded in the JIT compiler code.  Major changes of the
language are likely to correspond to major surgery of the JIT.

@@

Targetting platforms that are very different from C/Posix is work in
progress, but given that many of the initial components are shared with
the existing stack of transformations leading to C, we are confident
that this work will soon give results.  Moreover, we believe that these
results will show reasonable efficiency, because the back-ends for VMs
like Squeak and .NET can take advantage of high-level input (as opposed
to trying to translate, say, C-like code to Smalltalk).
